On the relation of the entanglement spectrum to the bulk polarization \\ DRAFT

### 2020-11-03 version2/draft.tex
Final lic version
Changed some notation

Consider a generic, quadratic Hamiltonian in one dimension
\begin{equation}\label{eq:quadr_Ham}
\mathcal{H} = \sum_{ij,\alpha\beta} c_{i\alpha}^\dagger H_{ij,\alpha \beta}c_{j\beta}.
\end{equation}
We can obtain its single-particle eigenstates as
\begin{align}
\sum_{j\beta}H_{ij,\alpha\beta} u_{p\mu}^{j\beta} = E_{p\mu} u^{i\alpha}_{p\mu},
\end{align}
where $[U]_{j\beta,p\mu} = u_{p\mu}^{j\beta}$ is the unitary matrix that diagonalizes $H$. In translational invariant systems $u_{k\mu}^{j\beta} = e^{ikj}\tilde{u}_{k \mu}^{\beta}$, where $\tilde{u}_{k \mu}^{\beta}$ are the components of the eigenstates of the Bloch Hamiltonian, $\ket{\tilde{u}_{k\mu}}$ and $k$ is the momentum. One can now compute the Zak phase \cite{Zak1989}, which for a two-band  model is defined as the geometric phase acquired by the occupied state $\ket{\tilde{u}_k}$ as it winds around the Brillouin zone,
\begin{equation}
\gamma = \int_{0}^{2\pi} {\rm d}k\, i A_k, 
\label{eq:zak_phase}
\end{equation}
where $A_k = \bra{\tilde{u}_k}\partial_k \ket{\tilde{u}_k}$ is the Berry connection \cite{Berry1984}. 
The generalization to higher-band models is straightforward.

@@

The derivative of the polarization with respect to $\lambda$ was obtained \cite{KingSmith1993} as
\begin{equation}\label{eq:Bloch_polarization}
\partial_\lambda \tilde{\mathcal{P}}^{\rm Bloch} = \int_{0}^{2\pi} \frac{dk}{2\pi} \, \Omega_{\lambda k},
\end{equation}
which allows us to make the identification $ \tilde{\mathcal{P}}^{\rm Bloch} = \gamma/2\pi$. 
As discussed above,  this polarization is defined only modulo 1.
Its changes, however, are defined in $\mathbb{R}$. The polarization itself is not a physical observable. Only its changes, related to the current flowing through the system, are.

One can also compute the geometric phase obtained by threading a $U(1)$ flux through the ring \cite{Watanabe2018}, which allows us to compute a polarization in the absence of translational invariance. If the flux is introduced homogeneously through a vector potential $A_x = \Phi/L$ another polarization can be defined as
\begin{equation}
\mathcal{P} = \int_0^{2\pi}\frac{d\Phi}{2\pi} i\bra{\Psi^\Phi}\partial_\Phi \ket{\Psi^\Phi} + \frac{1}{2\pi}\Im\ln \bra{\Psi^0}e^{2\pi i \hat{P}} \ket{\Psi^{2\pi}}, 
\end{equation}
where $\ket{\Psi^{\Phi}}$ is the ground state in the presence of flux and $\hat{P}=\frac{1}{L}\sum_{j\alpha} j\hat{n}_{j\alpha}$ is the polarization operator, where $\hat{n}_{j\alpha}$ is the number operator. This is equivalent to the polarization obtained by Resta \cite{Resta1997} as
\begin{align}
\mathcal{P} = \frac{1}{2\pi} \Im \,\ln\bra{\Psi^{\Phi=0}}e^{2\pi i \hat{P}}\ket{\Psi^{\Phi=0}}.
\end{align}
This expression is particularly useful because it can be easily expressed in terms of single-particle eigenstates as
\begin{align}
\mathcal{P} = \frac{1}{2\pi}\Im \ln \, {\rm det }\!' \, S,
\label{eq:polarization_resta}
\end{align}
where the matrix $S$ is given by
\begin{equation}
S_{p\mu,q\nu} = \sum_{j\alpha} u_{p\mu}^{j \alpha \, \ast} e^{i\frac{2\pi}{L}j}u_{q\nu}^{j \alpha},
\end{equation}
and  ${\rm det }\!' $ indicates that the determinant is restricted to the space of \emph{occupied} single-particle states. Below we will compare the bulk polarization obtained using our method to the one obtained using equation~\eqref{eq:polarization_resta}.

There is one last bulk polarization we must consider. 
It is obtained when the flux is introduced in the bond between sites $j=1$ and $j=L$ via twisted boundary conditions. This is equivalent to performing the gauge transformation
\begin{align}\label{eq:Psi_tilde}
\ket{\tilde{\Psi}^\Phi}=e^{i\Phi \hat{P}}\ket{\Psi^\Phi},
\end{align}
which makes $\ket{\tilde{\Psi}^\Phi}$ fully periodic in $\Phi$. We can now define the bulk polarization
\begin{equation}
\tilde{\mathcal{P}} = \int_0^{2\pi}\frac{d\Phi}{2\pi} i\bra{\tilde{\Psi}^\Phi}\partial_\Phi \ket{\tilde{\Psi}^\Phi},
\label{eq:p_tilde}
\end{equation}
which is related to the current flowing through this bond \cite{Watanabe2018}. This bulk polarization is related to the previous one by 
\begin{equation}
\mathcal{P}= \tilde{\mathcal{P}}+\tilde{\mathcal{P}}_0,
\end{equation} 
where
\begin{equation}
\tilde {\mathcal{P}}_0= \int_0^{2\pi}\frac{d\Phi}{2\pi} \bra{\tilde{\Psi}^\Phi}\hat{P}\ket{\tilde{\Psi}^\Phi}.
\end{equation} 
%The previous discussion about the Zak phase being defined only modulo $2\pi$ although its changes can be defined as a real number applies in the same way to the polarization obtained in \eqref{eq:p_tilde}.
Similar to the discussion about the Bloch polarization below Eq.~\eqref{eq:Bloch_polarization}, both $\mathcal{P}$ and $\tilde{\mathcal{P}}$ are defined modulo 1, while their changes are defined in $\mathbb{R}$. For translationally invariant systems one finds that $\tilde{\mathcal{P}} = \tilde{\mathcal{P}}^{\rm Bloch}$ \cite{Watanabe2018}.

@@

\section{Alternative expression for the correlation matrix}\label{app:CM}
In this appendix, we show how to express the correlation matrix in terms of the Hamiltonian, used in Eq.~\eqref{eq:corr_mat2}. 
We consider the generic, quadratic Hamiltonian in one dimension of Eq.~\eqref{eq:quadr_Ham}, which is diagonalized by a unitary matrix $U$ with 
	\begin{align}\label{eq:D}
	H&=UDU^\dagger& \mbox{with } D=\mbox{diag}(E_{p\mu}).
	\end{align} 
In terms of the fermionic operators that diagonalize the Hamiltonian, 
\begin{align}
& \gamma_{i\alpha} = \sum_{j\beta}u_{j\beta}^{i\alpha \, \ast} c_{j\beta},
%& c_{i\alpha} = \sum_{j\beta} u^{j\beta}_{i\alpha} \gamma_{j\beta}
\end{align}
we find that the correlation matrix can be written as 
\begin{align}\label{eq:corr_mat}
C_{ij}^{\alpha \beta} =& \sum_{pq,\mu\nu} u^{q\nu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} \expval{\gamma^\dagger_{p\mu} \gamma_{q\nu} }\nonumber \\
=&  \sum_{p\mu} u^{p\mu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} \expval{\gamma^\dagger_{p\mu} \gamma_{p\mu} }\nonumber \\
=&  \sum_{p\mu} u^{p\mu }_{j\beta} u^{p\mu \, \ast}_{i\alpha} [1-{\rm sign}(E_{p\mu})]/2.
\end{align}
The first term of the last line is simply a kronecker delta between both sets of indices. The second term can be rewritten, using Eq.~\eqref{eq:D}, as
\begin{align}
UD(\abs{D})^{-1}U^\dagger =& UDU^\dagger U(D^2)^{-1/2}U^\dagger
\end{align}
We can rewrite this expression further by noting that 
\begin{align}
[U (D^2)^{-1/2} U^\dagger]^2 =& U (D^2)^{-1/2} U^\dagger U  (D^2)^{-1/2} U^\dagger \nonumber\\
=&  U (D^2)^{-1/2}  (D^2)^{-1/2} U^\dagger \nonumber\\
=&  U (D^2)^{-1} U^\dagger \nonumber\\
=&  (U D^2 U^\dagger)^{-1}.
\end{align}
Therefore,
\begin{align*}
U (D^2)^{-1/2} U^\dagger =&(U D^2 U^\dagger)^{-1/2} 
\end{align*}
and we conclude that
\begin{align}\label{eq:UDDU}
UD(\abs{D})^{-1}U^\dagger =& UDU^\dagger(U D^2 U^\dagger)^{-1/2} \nonumber\\ 
=& H(H^2)^{-1/2}.
\end{align}
Combining Eq.s~\eqref{eq:corr_mat},  and \eqref{eq:UDDU}, we arrive at the final expression of the correlation matrix in Eq.~\eqref{eq:corr_mat2}.

>>>>

Consider a generic, quadratic Hamiltonian in one dimension
\begin{equation}\label{eq:quadr_Ham}
\mathcal{H} = \sum_{ij,\alpha\beta} c_{i\alpha}^\dagger H_{ij,\alpha \beta}c_{j\beta}.
\end{equation}
We can obtain its single-particle eigenstates as
\begin{align}
\sum_{j\beta}H_{ij,\alpha\beta} \psi_{p\mu}^{j\beta} = E_{p\mu} \psi^{i\alpha}_{p\mu},
\end{align}
where $[U]_{j\beta,p\mu} = \psi_{p\mu}^{j\beta}$ is the unitary matrix that diagonalizes $H$. In translational invariant systems $\psi_{k\mu}^{j\beta} = e^{ikj}u_{k \mu}^{\beta}$, where $u_{k \mu}^{\beta}$ are the components of the eigenstates of the Bloch Hamiltonian, $\ket{u_{k\mu}}$ and $k$ is the momentum. For simplicity we consider a unit cell with a single site. One can now compute the Zak phase \cite{Zak1989}, which for a two-band  model is defined as the geometric phase acquired by the occupied state $\ket{u_k}$ as it winds around the Brillouin zone,
\begin{equation}
\gamma = \int_{0}^{2\pi} {\rm d}k\, i A_k, 
\label{eq:zak_phase}
\end{equation}
where $A_k = \bra{u_k}\partial_k \ket{u_k}$ is the Berry connection \cite{Berry1984}. 
The generalization to higher-band models is straightforward.

@@

The derivative of the polarization with respect to $\lambda$ was obtained \cite{KingSmith1993} as
\begin{equation}\label{eq:Bloch_polarization}
\partial_\lambda \mathcal{P}^{\rm Bloch} = \int_{0}^{2\pi} \frac{dk}{2\pi} \, \Omega_{\lambda k},
\end{equation}
which allows us to make the identification $ \mathcal{P}^{\rm Bloch} = \gamma/2\pi$. 
As discussed above,  this polarization is defined only modulo 1.
Its changes, however, are defined in $\mathbb{R}$. The polarization itself is not a physical observable. Only its changes, related to the current flowing through the system, are.

One can also compute the geometric phase obtained by threading a $U(1)$ flux through the ring \cite{Watanabe2018}, which allows us to compute a polarization in the absence of translational invariance. If the flux is introduced homogeneously through a vector potential $A_x = \Phi/L$ another polarization can be defined as
\begin{equation}
\mathcal{P} = \int_0^{2\pi}\frac{d\Phi}{2\pi} i\bra{\Psi^\Phi}\partial_\Phi \ket{\Psi^\Phi} + \frac{1}{2\pi}\Im\ln \bra{\Psi^0}e^{2\pi i \hat{P}} \ket{\Psi^{2\pi}}, 
\end{equation}
where $\ket{\Psi^{\Phi}}$ is the ground state in the presence of flux and $\hat{P}=\frac{1}{L}\sum_{j\alpha} j\hat{n}_{j\alpha}$ is the polarization operator, where $\hat{n}_{j\alpha}$ is the number operator. This is equivalent to the polarization obtained by Resta \cite{Resta1997} as
\begin{align}
\mathcal{P} = \frac{1}{2\pi} \Im \,\ln\bra{\Psi^{\Phi=0}}e^{2\pi i \hat{P}}\ket{\Psi^{\Phi=0}}.
\end{align}
This expression is particularly useful because it can be easily expressed in terms of single-particle eigenstates as
\begin{align}
\mathcal{P} = \frac{1}{2\pi}\Im \ln \, {\rm det }\!' \, S,
\label{eq:polarization_resta}
\end{align}
where the matrix $S$ is given by
\begin{equation}
S_{p\mu,q\nu} = \sum_{j\alpha} \psi_{p\mu}^{j \alpha \, \ast} e^{i\frac{2\pi}{L}j}\psi_{q\nu}^{j \alpha},
\end{equation}
and  ${\rm det }\!' $ indicates that the determinant is restricted to the space of \emph{occupied} single-particle states. Below we will compare the bulk polarization obtained using our method to the one obtained using equation~\eqref{eq:polarization_resta}.

There is one last bulk polarization we must consider. 
It is obtained when the flux is introduced in the bond between sites $j=1$ and $j=L$ via twisted boundary conditions. This is equivalent to performing the gauge transformation
\begin{align}\label{eq:Psi_tilde}
\ket{\tilde{\Psi}^\Phi}=e^{i\Phi \hat{P}}\ket{\Psi^\Phi},
\end{align}
which makes $\ket{\tilde{\Psi}^\Phi}$ fully periodic in $\Phi$. We can now define the bulk polarization
\begin{equation}
\tilde{\mathcal{P}} = \int_0^{2\pi}\frac{d\Phi}{2\pi} i\bra{\tilde{\Psi}^\Phi}\partial_\Phi \ket{\tilde{\Psi}^\Phi},
\label{eq:p_tilde}
\end{equation}
which is related to the current flowing through this bond \cite{Watanabe2018}. This bulk polarization is related to the previous one by 
\begin{equation}
\mathcal{P}= \tilde{\mathcal{P}}+\tilde{\mathcal{P}}_0,
\end{equation} 
where
\begin{equation}
\tilde {\mathcal{P}}_0= \int_0^{2\pi}\frac{d\Phi}{2\pi} \bra{\tilde{\Psi}^\Phi}\hat{P}\ket{\tilde{\Psi}^\Phi}.
\end{equation} 
%The previous discussion about the Zak phase being defined only modulo $2\pi$ although its changes can be defined as a real number applies in the same way to the polarization obtained in \eqref{eq:p_tilde}.
Similar to the discussion about the Bloch polarization below Eq.~\eqref{eq:Bloch_polarization}, both $\mathcal{P}$ and $\tilde{\mathcal{P}}$ are defined modulo 1, while their changes are defined in $\mathbb{R}$. For translationally invariant systems one finds that $\mathcal{P} = \mathcal{P}^{\rm Bloch}+\frac{L-1}{2}\nu$, here $\nu$ is the number of occupied bands \cite{Watanabe2018}.

@@

\section{Alternative expression for the correlation matrix}\label{app:CM}
In this appendix, we show how to express the correlation matrix in terms of the Hamiltonian, used in Eq.~\eqref{eq:corr_mat2}. 
We consider the generic, quadratic Hamiltonian in one dimension of Eq.~\eqref{eq:quadr_Ham}, which is diagonalized by a unitary matrix $U$ with 
	\begin{align}\label{eq:D}
	H&=UDU^\dagger& \mbox{with } D=\mbox{diag}(E_{p\mu}).
	\end{align} 
In terms of the fermionic operators that diagonalize the Hamiltonian, 
\begin{align}
& \gamma_{i\alpha} = \sum_{j\beta}\psi_{j\beta}^{i\alpha \, \ast} c_{j\beta},
%& c_{i\alpha} = \sum_{j\beta} \psi^{j\beta}_{i\alpha} \gamma_{j\beta}
\end{align}
we find that the correlation matrix can be written as 
\begin{align}\label{eq:corr_mat}
C_{ij}^{\alpha \beta} =& \sum_{pq,\mu\nu} \psi^{q\nu }_{j\beta} \psi^{p\mu \, \ast}_{i\alpha} \expval{\gamma^\dagger_{p\mu} \gamma_{q\nu} }\nonumber \\
=&  \sum_{p\mu} \psi^{p\mu }_{j\beta} \psi^{p\mu \, \ast}_{i\alpha} \expval{\gamma^\dagger_{p\mu} \gamma_{p\mu} }\nonumber \\
=&  \sum_{p\mu} \psi^{p\mu }_{j\beta} \psi^{p\mu \, \ast}_{i\alpha} [1-{\rm sign}(E_{p\mu})]/2.
\end{align}
The first term of the last line is simply a kronecker delta between both sets of indices. The second term can be rewritten, using Eq.~\eqref{eq:D}, as
\begin{align}
UD(\abs{D})^{-1}U^\dagger =& UDU^\dagger U(D^2)^{-1/2}U^\dagger
\end{align}
We can rewrite this expression further by noting that 
\begin{align}
[U (D^2)^{-1/2} U^\dagger]^2 =& U (D^2)^{-1/2} U^\dagger U  (D^2)^{-1/2} U^\dagger \nonumber\\
=&  U (D^2)^{-1/2}  (D^2)^{-1/2} U^\dagger \nonumber\\
=&  U (D^2)^{-1} U^\dagger \nonumber\\
=&  (U D^2 U^\dagger)^{-1}.
\end{align}
Therefore,
\begin{align*}
U (D^2)^{-1/2} U^\dagger =&(U D^2 U^\dagger)^{-1/2} 
\end{align*}
and we conclude that
\begin{align}\label{eq:UDDU}
UD(\abs{D})^{-1}U^\dagger =& UDU^\dagger(U D^2 U^\dagger)^{-1/2} \nonumber\\ 
=& H(H^2)^{-1/2}.
\end{align}
Combining Eq.s~\eqref{eq:corr_mat},  and \eqref{eq:UDDU}, we arrive at the final expression of the correlation matrix in Eq.~\eqref{eq:corr_mat2}.
